---
title: "Homework 2 - P8105"
output: github_document
author: Grace Santos 
date: "2023-09-27"
---


```{r}
library(tidyverse)
```


## Problem 1 

```{r, pols_month}
 month_df = 
    tibble(
    month_num = 1:12, 
    month_abb = month.abb, 
    month = month.name)

pols_month_df = 
  read_csv("fte_data/pols-month.csv") |> 
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |> 
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez"))
```

```{r, snp}
snp_df = 
  read_csv("fte_data/snp.csv",
           col_types = cols(date = col_date(format = "%m/%d/%y"))) |> 
  separate(date, into = c("year", "month_num", "day"), convert = TRUE) |> 
  mutate(
    year = if_else(year > 2023, year - 100, year)) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, close)
```

```{r, unemployment}
unemploy_df = 
  read_csv("fte_data/unemployment.csv") |> 
  rename(year = Year) |> 
  pivot_longer(
    Jan:Dec,
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```

```{r}
fte_data = 
  left_join(pols_month_df, snp_df) |> 
  left_join(x = _, y = unemploy_df)

str(fte_data)
```

When looking at the final joined data, there are `NA` values in the `close` and `unemployment` variables, which indicates missing values. 

The `pols_month_df` data has `r nrow(pols_month_df)` observations and `r ncol(pols_month_df)` variables and tells us about the party affiliation distribution (democrat or republican) for governors and senators for a given year from years `r pols_month_df |> pull(year) |> min()` to `r pols_month_df |> pull(year) |> max()`. It also tells us whether the sitting president was a democrat or republican. 

The `snp_df` data has `r nrow(snp_df)` observations and `r ncol(snp_df)` variables, ranging from years `r snp_df |> pull(year) |> min()` to `r snp_df |> pull(year) |> max()`. 

The `unemploy_df` data has `r nrow(unemploy_df)` observations and `r ncol(unemploy_df)` variables ranging from years `r unemploy_df |> pull(year) |> min()` to `r unemploy_df |> pull(year) |> max()`. 

In Januarys in or after 1975 in which a democrat was president, the **average unemployment rate was `r filter(fte_data, month == "January", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`**.  The average unemployment rate over the same time period in which a republican was president was `r filter(fte_data, month == "January", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.


## Problem 2 
```{r}
mr_trash = 
  readxl::read_excel("trash_data/202207 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:M549") |> 
  janitor::clean_names() |> 
  mutate(
    year = as.numeric(year), 
    name = "mr_trash_wheel",
    homes_powered = (weight_tons *500)/30)

prof_trash = 
    readxl::read_excel("trash_data/202207 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:L96") |> 
  janitor::clean_names() |> 
  mutate(
    name = "prof_trash_wheel",
    homes_powered = (weight_tons *500)/30)

gwy_trash = 
  readxl::read_excel("trash_data/202207 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", range = "A2:J108") |> 
  janitor::clean_names() |>
  mutate(
    name = "gwynnda_trash_wheel",
    homes_powered = (weight_tons *500)/30)

trash_tidy = 
  bind_rows(mr_trash, prof_trash, gwy_trash) |> 
  select(name, everything())
```

Need to still write paragraph - want clarity on code before finishing the problem 

## Problem 3 
```{r}
bl_data = 
  read_csv("mci_data/MCI_baseline.csv", skip = 1, col_names = TRUE) |> 
  janitor::clean_names() |> 
  mutate(
    sex = 
      case_match(
        sex, 
        0 ~ "female",
        1 ~ "male"),
    sex = as.factor(sex),
    apoe4 = 
      case_match(
        apoe4, 
        0 ~ "non-carrier",
        1 ~ "carrier"),
      apoe4 = as.factor(apoe4))

amy_data = 
  read_csv("mci_data/mci_amyloid.csv", skip = 1, col_names = TRUE) |> 
  janitor::clean_names() |> 
  rename("id" = "study_id")

mci_tidy = 
 full_join(bl_data, amy_data, by = "id")
```

